Parallel Adaptive Importance Sampling

### 2016-06-02 PAIS.tex
Abstract/Intro/Prelims
Updates to the first two sections and the abstract to reflect changes
in the rest of the document.

\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e: simon.cotter@manchester.ac.uk} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Parallel Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
  Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. As applications of these methods increase in size and complexity, the need for efficient
  methods which can exploit the parallel architectures which are
  prevalent in high performance computing increases. In this paper, we
  aim to develop a framework for scalable parallel MCMC algorithms. At
  each iteration, an importance sampling proposal distribution is
  formed using the current states of all of the chains within an
  ensemble. Once weighted samples have been produced from this, a
  state-of-the-art resampling method is then used to create an evenly
  weighted sample ready for the next iteration. We demonstrate that
  this parallel adaptive importance sampling (PAIS) method outperforms
  naive parallelisation of serial MCMC methods using the same number
  of processors, for low dimensional problems, and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of processors used.
\end{abstract}
\begin{keywords}MCMC, parallel, importance sampling, Bayesian, inverse problems.
\end{keywords}
\section{Introduction}
%\subsection{Lit. review}
%\subsection{Optimal transport resampler}
%\subsection{Function Space MCMC}
Markov chain Monte Carlo (MCMC) methods are a powerful family of tools
that allow us to sample from complex probability distributions. MCMC
methods were first developed in the 70s\cite{hastings1970monte}, and
with the development of and with the development of faster more
powerful computers, have become ever more important in a whole range
of fields in statistics, science and engineering.  In particular, when
considering Bayesian inverse problems, each MCMC step may involve the
numerical solution of one or more PDE. As many samples are usually
required before Monte Carlo error is reduced to acceptable levels,
these types of problem remain frustratingly out of our grasp.

@@

In this paper, we present a framework for parallelisation of
importance sampling, which can be built around any of the current
Metropolis-based methodologies in order to create an efficient target
proposal from the current state of all of the chains in the
ensemble. The idea is to consider the parallel chains as an ensemble,
and to resample using a transformation based on optimal transport.
Samplers based on optimal transport have also been considered in
\cite{el2012bayesian}.

In Section \ref{Sec:Prelim} we outline some preliminaries,
including the general set up for Bayesian inverse problems, the
preconditioned Crank-Nicolson Langevin (pCNL) algorithm and a
brief review of the optimal transport resampler,  both of which we will be
employing within the algorithm. We describe the Parallel Adaptive
Importance Sampler (PAIS) in Section \ref{Sec:PAIS}, and in Section
\ref{Sec:adapt} we describe how algorithmic parameters can be
automatically tuned to provide optimal convergence. In
Section \ref{Sec:Num}, we present some numerical experiments that
demonstrate the savings available by employing this approach as
opposed to a naive/trivial parallelisation of existing MCMC
methods. In Section \ref{Sec:Conc}, we summarise our results and
suggest some areas for future investigation.

@@

\subsection{The preconditioned Crank-Nicolson Langevin (pCNL) algorithm}\label{Sec:pCNL}
In recent years, work has been carried out to frame MCMC proposal
distributions on function space\cite{cotter2013mcmc}. These new
discretisations perform comparably with the original versions in low
dimensions. If gradient information regarding the observation operator is
available, then a range of MCMC methods are available which exploit this information to improve mixing rates. One
example of such an algorithm is MALA. In \cite{cotter2013mcmc}, a function space version of this
method was presented, the pCNL algorithm, and is described in full in Table
\ref{tab:pCNL}. The proposal used in this method comes about through
a Crank-Nicolson approximation of the Langevin SDE, whose invariant
measure is the posterior measure $\mu_Y$.

\begin{table}
\begin{mdframed}
\begin{algorithmic}
\STATE $X = x_0$
\FOR{$i=1,2,3,\ldots$}
\STATE $Y = (2+\delta)^{-1}\left[(2 - \delta)X_{i-1}-
2\delta\mathcal{C}\nabla \Phi(u)+
\sqrt{8\delta} W\right] $, $W \sim \mu_0$
\STATE $a(X_{i-1},Y) = \min \left \{ 1,  \exp(\Phi(X_{i-1}) - \Phi(Y) ) \right \}$.
\STATE $u \sim U([0,1])$
\IF{$u<a(X_{i-1},Y)$}
\STATE $X_i = Y$
\ELSE 
\STATE $X_i = X_{i-1}$
\ENDIF
\ENDFOR
\end{algorithmic}
\end{mdframed}\caption{A pseudo-code representation of the preconditioned Crank-Nicolson Langevin
   (pCNL) algorithm. $\delta \in (0,2]$ is a step size parameter.}
\label{tab:pCNL}
\end{table}

@@

One advantage of the particle filter is that there are convergence
results for this method as the number of particles is increased. The
downside is that, the effective sample size decreases at each iteration. One way to tackle
this is to employ a resampling scheme. The aim of a successful
resample is to take your unevenly weighted ensemble and return a new
ensemble of particles with even weights which is highly correlated to the original samples.

@@

A Monte Carlo algorithm can be implemented to resample from a weighted ensemble. We create a weighted sample, then solve the optimal transport problem which produces the coupling described above, we can draw a new sample from the evenly weighted distribution. Reich suggests using the mean of the evenly weighted distribution to produce a consistent estimator.

Analysis of this method shows that as the ensemble size increases, the statistics of the evenly weighted sample approach those of the posterior distribution, at least well enough for a proposal distribution as described in Section~\ref{Sec:PAIS}. The histogram of the evenly weighted sample exhibits small oscillations in the tails of the posterior, and also struggles to deal with discontinuities.

\subsection{Deficiencies of Metropolis-type MCMC schemes}
All MCMC methods are trivially parallelisable. One can take a method
and simply implement it simultaneously over a set of processes. All
of the states of all of the processes can be recorded, and in the time
that it takes one process to draw $N$ samples, $M$ processes can draw
$NM$ samples.

However, we argue that this is a far from optimal
scenario. First of all, unless we have a lot of information about the posterior, we will begin the algorithm a long way from statistical equilibrium. Some initial iterations then are not samples from the posterior, and must be thrown away. This process is known as the burn-in. In a trivially parallelised scenario, each process must perform this process independently.

Moreover, many MCMC algorithms suffer from poor mixing, especially in
multimodal systems. The time for an MCMC trajectory to switch between modes can be
large and given that a large number of switches are required before we have a
good idea of the relative probability densities of these different
regions, it can be prohibitively expensive.

@@

These deficiencies of the trivial method of parallelising MCMC methods
motivated the development of the Parallel Adaptive Importance Sampler
(PAIS). In the next section we will introduce the method in its most
general form. We will then introduce the version that we have
implemented, which utilises the resampler recently suggested by
Reich\cite{reich2013nonparametric} described in Section
\ref{sec:filters}, and the pCNL proposal distribution described in
Section \ref{Sec:pCNL}.

>>>>

\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e: simon.cotter@manchester.ac.uk} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Parallel Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
  Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. As applications of these methods increase in size and complexity, the need for efficient
  methods which can exploit the parallel architectures which are
  prevalent in high performance computing increases. In this paper, we
  aim to develop a framework for scalable parallel MCMC algorithms. At
  each iteration, an importance sampling proposal distribution is
  formed using the current states of all of the chains within an
  ensemble. Once weighted samples have been produced from this, a
  state-of-the-art resampling method is then used to create an evenly
  weighted sample ready for the next iteration. We demonstrate that
  this parallel adaptive importance sampling (PAIS) method outperforms
  naive parallelisation of serial MCMC methods using the same number
  of processors, for low dimensional problems, and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of processors used. We also introduce a new resampling
  strategy, approximate multinomial resampling (AMR), which while not
  as accurate as other schemes is substantially costly for large
  ensemble sizes, which can then be used in conjunction with PAIS for
  complex problems.
\end{abstract}
\begin{keywords}MCMC, parallel, importance sampling, Bayesian, inverse problems.
\end{keywords}
\section{Introduction}
%\subsection{Lit. review}
%\subsection{Optimal transport resampler}
%\subsection{Function Space MCMC}
Markov chain Monte Carlo (MCMC) methods are a powerful family of tools
that allow us to sample from complex probability distributions. MCMC
methods were first developed in the 70s\cite{hastings1970monte}, and
with the development of faster, more
powerful computers, have become ever more important in a whole range
of fields in statistics, science and engineering.  In particular, when
considering Bayesian inverse problems, each MCMC step may involve the
numerical solution of one or more PDE. As many samples are usually
required before Monte Carlo error is reduced to acceptable levels,
these types of problem remain frustratingly out of our grasp.

@@

In this paper, we present a framework for parallelisation of
importance sampling, which can be built around many of the current
Metropolis-based methodologies in order to create an efficient target
proposal from the current state of all of the chains in the
ensemble. The idea is to consider the parallel chains as an ensemble,
and to resample using a transformation based on optimal transport.
Samplers based on optimal transport have also been considered in
\cite{el2012bayesian}.

***SECTIONS SUMMARY NEEDS REWRITING***

@@

% \subsection{The preconditioned Crank-Nicolson Langevin (pCNL) algorithm}\label{Sec:pCNL}
% In recent years, work has been carried out to frame MCMC proposal
% distributions on function space\cite{cotter2013mcmc}. These new
% discretisations perform comparably with the original versions in low
% dimensions. If gradient information regarding the observation operator is
% available, then a range of MCMC methods are available which exploit this information to improve mixing rates. One
% example of such an algorithm is MALA. In \cite{cotter2013mcmc}, a function space version of this
% method was presented, the pCNL algorithm, and is described in full in Table
% \ref{tab:pCNL}. The proposal used in this method comes about through
% a Crank-Nicolson approximation of the Langevin SDE, whose invariant
% measure is the posterior measure $\mu_Y$.

% \begin{table}
% \begin{mdframed}
% \begin{algorithmic}
% \STATE $X = x_0$
% \FOR{$i=1,2,3,\ldots$}
% \STATE $Y = (2+\delta)^{-1}\left[(2 - \delta)X_{i-1}-
% 2\delta\mathcal{C}\nabla \Phi(u)+
% \sqrt{8\delta} W\right] $, $W \sim \mu_0$
% \STATE $a(X_{i-1},Y) = \min \left \{ 1,  \exp(\Phi(X_{i-1}) - \Phi(Y) ) \right \}$.
% \STATE $u \sim U([0,1])$
% \IF{$u<a(X_{i-1},Y)$}
% \STATE $X_i = Y$
% \ELSE 
% \STATE $X_i = X_{i-1}$
% \ENDIF
% \ENDFOR
% \end{algorithmic}
% \end{mdframed}\caption{A pseudo-code representation of the preconditioned Crank-Nicolson Langevin
%    (pCNL) algorithm. $\delta \in (0,2]$ is a step size parameter.}
% \label{tab:pCNL}
% \end{table}

@@

One advantage of the particle filter is that there are convergence
results for this method as the number of particles is increased. One
downside is that the effective sample size decreases at each
iteration, resulting in degeneration of the approximation of the
posterior. One way to tackle
this is to employ a resampling scheme. The aim of a successful
resample is to take your unevenly weighted ensemble and return a new
ensemble of particles with even weights which is highly correlated to
the original samples. Another downside is that the required ensemble
size increases quickly with dimension, making it difficult to use in
high-dimensional problems.

@@

A Monte Carlo algorithm can be implemented to resample from a weighted
ensemble. We create a weighted sample, then solve the optimal
transport problem which produces the coupling described above, we can
draw a new sample from the evenly weighted distribution. Reich
suggests using the mean of the evenly weighted distribution to produce
a consistent estimator. Analysis of this method shows that as the
ensemble size increases, the statistics of the evenly weighted sample
approach those of the posterior distribution.

\subsection{Deficiencies of Metropolis-type MCMC schemes}
All MCMC methods are trivially parallelisable. One can take a method
and simply implement it simultaneously over a set of processors. All
of the states of all of the processes can be recorded, and in the time
that it takes one process to draw $N$ samples, $M$ processes can draw
$NM$ samples.

However, we argue that this is not an optimal
scenario. First of all, unless we have a lot of information about the
posterior, we will intialise the algorithm's initial state in the
tails of the distribution. The samples that are initially made as the
algorithm finds its way to the mode(s) of the distribution cannot be
considered to samples from the target distribution, and must be thrown
away. This process is known as the burn-in. In a trivially
parallelised scenario, each process must perform this process
independently, and therefore mass parallelisation makes no inroads to
cutting this cost.

Moreover, many MCMC algorithms suffer from poor mixing, especially in
multimodal systems. The amount of samples that it takes for an MCMC trajectory to switch between modes can be
large, and given that a large number of switches are required before we have a
good idea of the relative probability densities of these different
regions, it can be prohibitively expensive.

@@

These deficiencies of the trivial method of parallelising MCMC methods
motivated the development of the Parallel Adaptive Importance Sampler
(PAIS). In the next section we will introduce the method in its most
general form.
